Modal Abstractions for Virtualizing Memory Addresses

### 2023-11-16 examples.tex
condensing new p2v example

Work to verify the full page table walk is ongoing, but we have verified the key step for a small x86-64 kernel, which is the physical-to-virtual
conversion (often appearing as a macro \texttt{P2V} in C source code). In our small kernel, as in larger kernels, \texttt{P2V} is actually just addition,
though the Hoare triple we verify for this add instruction is rather complex:

\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ulcorner \textsf{KERNELBASE} + \paddr > 0 \urcorner }_{\rtv}$
$\specline{ \texttt{rip}\mapsto_{\textsf{r}} \textsf{iv} \ast \texttt{rflags}\mapsto_{\textsf{r}} \textsf{flaglist} \ast \texttt{rbx}\mapsto_{\textsf{r}} \paddr  }_{\rtv}$
$\specline{ \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$
;; This offsets the physical address bt KERNELBASE!
add rbx, KERNELBASE
$\specline{\exists \qfraczero; \qfrac,\textsf{ flaglistadd, w, flaglistadd, val}, \textsf{P} \ast \texttt{rip}\mapsto_{\textsf{r}} \textsf{iv+7} \ast \texttt{rflags}\mapsto_{\textsf{r}} \textsf{flaglistadd} \ast \texttt{rbx}\mapsto_{\textsf{r}} \paddr + \textsf{KERNELBASE} }_{\rtv}$
$\specline{\bigast{(\vaddr, \paddr)\in \ptablestore}{\exists\;(\textsf{l4e, l3e, l2e, l1e, } \paddr)\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr\textsf{, l4e, l3e, l2e, l1e,} \paddr)}}_{\rtv}$
$\specline{\mathcal{R} -\ast \bigast{(\paddr) \in \Xi}{\exists\; (\textsf{qfrac, q, val, va}) \ldotp \ulcorner \textsf{va} = \paddr + \textsf{KERNELBASE} \urcorner \ast \sumwalkabs\vaddr\qfrac\paddr  \ast \paddr \mapsto_{\mathsf{p}}\{\textsf{qfrac}\} \; \vale\ast  \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner}}_{\rtv}$
$\specline{ \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}} \ast \ownGhostpv\gammaPred{\authfull{\pvmapping\Xi}}}$
$\specline{ \sumapaces\rtv\delta  \ast\sumwalkabs\vaddr\qfrac\paddr \ast \sumpv\paddr\qfraczero\true \ast  \paddr \mapsto_{\mathsf{p}}\{\textsf{qfrac}\} \; \vale \ast \textsf{P}}_{\rtv}$
$\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\; \sqsubseteq $
$\specline{\exists \qfraczero; \qfrac,\textsf{ flaglistadd, w, flaglistadd, val}, \textsf{P} \ast \texttt{rip}\mapsto_{\textsf{r}} \textsf{iv+7} \ast \texttt{rflags}\mapsto_{\textsf{r}} \textsf{flaglistadd} \ast \texttt{rbx}\mapsto_{\textsf{r}} \paddr + \textsf{KERNELBASE} }_{\rtv}$
$\specline{\bigast{(\vaddr, \paddr)\in \ptablestore}{\exists\;(\textsf{l4e, l3e, l2e, l1e, } \paddr)\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr\textsf{, l4e, l3e, l2e, l1e,} \paddr)}}_{\rtv}$
$\specline{\mathcal{R} -\ast \bigast{(\paddr) \in \Xi}{\exists\; (\textsf{qfrac, q, val, va}) \ldotp \ulcorner \textsf{va} = \paddr + \textsf{KERNELBASE} \urcorner \ast \sumwalkabs\vaddr\qfrac\paddr  \ast \paddr \mapsto_{\mathsf{p}}\{\textsf{qfrac}\} \; \vale\ast  \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner}}_{\rtv}$
$\specline{ \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}} \ast \ownGhostpv\gammaPred{\authfull{\pvmapping\Xi}}}$
$\specline{ \sumapaces\rtv\delta \ast \sumpv\paddr\qfraczero\true \ast \textsf{P}}_{\rtv}$
$\specline{\vaddr \mapsto_{\texttt{vpte}} \{\textsf{q}, \textsf{qfrac}\} \paddr \;(\texttt{val})}$
\end{lstlisting}
\caption{Identity mapping of a physical address to a virtual one.}
\label{fig:mapping_code}
\end{figure}

We can see that this specification neatly encapsulates reasoning about invariants on the page tables in terms of general modal assumptions.\todo{speculative
but ought to turn out this way}
This is a simple piece of code whose functionality is critical and whose correctness is highly non-trivial. No prior work engages with this problem.

>>>>

\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ulcorner \textsf{KERNELBASE} + \paddr > 0 \urcorner \ast \texttt{rbx}\mapsto_{\textsf{r}} \paddr \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$
add rbx, KERNELBASE ;; shift physical address by identity mapping constant
$\specline{\exists \qfraczero; \qfrac,\textsf{ w, val}, \textsf{P} \ast \texttt{rbx}\mapsto_{\textsf{r}} \paddr + \textsf{KERNELBASE} }_{\rtv}$
$\specline{\bigast{(\vaddr, \paddr)\in \ptablestore}{\exists\;(\textsf{l4e, l3e, l2e, l1e, } \paddr)\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr\textsf{, l4e, l3e, l2e, l1e,} \paddr)}}_{\rtv}$
$\specline{\mathcal{R} -\ast \bigast{(\paddr) \in \Xi}{\exists\; (\textsf{qfrac, q, val, va}) \ldotp \ulcorner \textsf{va} = \paddr + \textsf{KERNELBASE} \urcorner \ast \sumwalkabs\vaddr\qfrac\paddr  \ast \paddr \mapsto_{\mathsf{p}}\{\textsf{qfrac}\} \; \vale\ast  \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner}}_{\rtv}$
$\specline{ \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}} \ast \ownGhostpv\gammaPred{\authfull{\pvmapping\Xi}}}$
$\specline{ \sumapaces\rtv\delta \ast \sumpv\paddr\qfraczero\true \ast \textsf{P}}_{\rtv}$
$\specline{\vaddr \mapsto_{\texttt{vpte}} \{\textsf{q}, \textsf{qfrac}\} \paddr \;(\texttt{val})}$
\end{lstlisting}
\todo[inline]{Ismail: Should \delta be something else? $\delta$ was used for $\theta$, we need a different gname for $\Xi$.}
\caption{Converting a physical address of a PTE to a virtual address (without instruction pointer or flag updates).}
\label{fig:p2v}
\end{figure}

Work to verify the full page table walk is ongoing, but we have verified the key step for a small x86-64 kernel, which is the physical-to-virtual
conversion (often appearing as a macro \texttt{P2V} in C source code). In our small kernel, as in larger kernels, \texttt{P2V} is actually just addition,
though the Hoare triple we verify for this add instruction is rather complex, with the specification shown in Figure \ref{fig:p2v}.
Beyond a frame ($P$), the precondition requires the current address space invariant, a check that the addition will not overflow, the physical address to convert
(in \lstinline|rbx|), knowledge that the current address space is valid, and knowledge that the physical address is in a range dedicated
to page table entries (as this physical-to-virtual conversion is only guaranteed for page table memory). The latter uses an additional ghost set added to the per-address-space
invariant per Figure \ref{fig:peraspaceinvariant_with_p2v_extension}. The extended invariant owns, for each memory location dedicated to page tables, some fraction of
that physical memory, and also guarantees the offset is the correct calculation.\todo{Colin needs a second pass here}
This is a simple piece of code whose functionality is critical and whose correctness is highly non-trivial. No prior work engages with this problem.
